A Spectral Analysis of the Domain Decomposed Adjoint Neumann-Ulam Method

### 2012-11-21 doc/mandc2013/notes/spectral_theory.tex
finished theory section. experiment results next

  The domain decomposed behavior of the adjoint Neumann-Ulam method
  for solving linear operator equations is analyzed using the spectral
  properties of the linear operator. Relationships for the average
  length of the adjoint random walks and the fraction of histories
  leaking from a domain are made with respect to the eigenvalues of
  the linear operator. The one-speed, two-dimensional neutron
  diffusion equation is used as a model problem to test the spectral
  theory for symmetric operators. These relationships will serve as
  guidelines for selecting an appropriate parallel algorithm strategy
  and provide a basis for performance models for future parallel
  implementations of the adjoint Neumann-Ulam method.

@@

For reactor physics Monte Carlo simulations, domain decomposition has
been identified as a key principle in moving forward in high
performance computing
\cite{brunner_comparison_2006,siegel_analysis_2012}. To accomplish
this, we recognize from the literature that stochastic histories must
be transported from domain to domain as the simulation progresses and
they transition to states that are not in the local domain. Because we
have chosen a domain decomposition strategy in a parallel environment,
this means that communication of these histories must occur between
compute nodes owning neighboring pieces of the global domain. We wish
to characterize this communication not only because communication is
in general expensive, but also because nearest-neighbor communication
sequences have poor algorithmic strong scaling \cite{gropp_cfd_2001}.
We observe this because nearest neighbor communication patterns scale
with the volume and surface sizes of the local domains. The larger the
volume, the more compute time is spent within the local volume and
therefore at some point you are limited by the latency of the system,
or the time it takes to receive the data to operate on. However, at a
certain point, when the surface of those volumes becomes large enough,
the amount of data that must be communicated grows large enough that
it may overcome the bandwidth limitations of the system, resulting in
a decrease in performance.

@@

In the adjoint Neumann-Ulam method, $k$ iterations, equivalent to $k$
applications of the iteration matrix, is effectively approximated by a
random walk of length $k$ to yield the summation in
Eq~(\ref{eq:adjoint_neumann_solution}). This random walk length, or
the number of transitions before the termination of a history (either
by the weight cutoff, absorption, or exiting the domain) is therefore
bound to the number of stationary iterations required to converge to
the specified tolerance. In the case of the adjoint Neumann-Ulam
method, no such tolerance exists, however, we have specified a weight
cutoff $W_c$ under which low-weight histories will be prematurely
terminated as their contributions are deemed minute. After $k$
iterations, a stationary method is terminated as the error has reached
some fraction, $\epsilon$, of the initial error:
\begin{equation}
  ||\ve{e}^{k}||_2 = \epsilon ||\ve{e}^0||_2\:.
  \label{eq:linear_k_iter_norm3}
\end{equation}
Per Eq~(\ref{eq:linear_k_iter_norm3}), we see that this fraction is
equivalent to $\epsilon = \rho(\ve{H})$. For the adjoint Neumann-Ulam
method, if we take this fraction to be the weight cutoff, a measure of
how accurately the contributions of a particular history to the
solution are tallied, we then have the following empirical relation
for $k$:
\begin{equation}
  k = \frac{ \log(W_c) }{ \log( \rho(\ve{H}) ) }\:.
  \label{eq:analytic_k}
\end{equation}
This then gives us a means to estimate the length of the random walks
derived from a particular linear operator based on the Eigenvalues of
its iteration matrix (independent of the linear operator splitting
chosen) and based on the weight cutoff parameter used in the
Neumann-Ulam method

\subsection{The Wigner Rational Approximation}
In a domain decomposed situtation, not all histories will remain
within the domain they started in and must instead be
communicated. This communication, expected to be expensive, was
analyzed by Siegel and colleagues for idealized, load balanced
situations for full-core Monte Carlo\cite{siegel_analysis_2012}.  To
quantify the number of particles that leak out of the local domain
they define a leakage fraction, $\lambda$, as:
\begin{equation}
  \lambda = \frac{average\ \#\ of\ particles\ leaving\ local\ domain}
          {total\ of\ \#\ of\ particles\ starting\ in\ local\ domain}\:.
  \label{eq:leakage_fraction}
\end{equation}
For their studies, they assumed that the value of $\lambda$ was
dependent on the total cross section via the Wigner rational
approximation. The approximation was developed to compute the neutron
leakage probability from by Wigner as a simple interpolation between
the small and large fuel lump cases.

@@

%%---------------------------------------------------------------------------%%
\section{Conclusion}

>>>>

  The domain decomposed behavior of the adjoint Neumann-Ulam method
  for solving linear operator equations is analyzed using the spectral
  properties of the linear operator. Relationships for the average
  length of the adjoint random walks and the fraction of histories
  leaking from a domain are made with respect to the Eigenvalues of
  the linear operator. In addition, a formal definition of the optical
  thickness of a linear operator is presented based in the spectral
  analysis. The one-speed, two-dimensional neutron diffusion equation
  is used as a model problem to test the spectral theory for symmetric
  operators. In general, the spectral theory shows good agreement with
  measured computational results.

@@

For reactor physics Monte Carlo simulations, domain decomposition has
been identified as a key principle in moving forward in high
performance computing
\cite{brunner_comparison_2006,siegel_analysis_2012}. To accomplish
this, we recognize from the literature that stochastic histories must
be transported from domain to domain as the simulation progresses and
they transition to states that are not in the local domain. Because we
have chosen a domain decomposition strategy in a parallel environment,
this means that communication of these histories must occur between
compute nodes owning neighboring pieces of the global domain. We wish
to characterize this communication not only because communication is
in general expensive, but also because nearest-neighbor communication
sequences have poor algorithmic strong scaling
\cite{gropp_high-performance_2001}.  We observe this because nearest
neighbor communication patterns scale with the volume and surface
sizes of the local domains. The larger the volume, the more compute
time is spent within the local volume and therefore at some point you
are limited by the latency of the system, or the time it takes to
receive the data to operate on. However, at a certain point, when the
surface of those volumes becomes large enough, the amount of data that
must be communicated grows large enough that it may overcome the
bandwidth limitations of the system, resulting in a decrease in
performance.

@@

In the adjoint Neumann-Ulam method, $k$ iterations, equivalent to $k$
applications of the iteration matrix, are approximated by a
random walk of length $k$ to yield the summation in
Eq~(\ref{eq:adjoint_neumann_solution}). This random walk length, or
the number of transitions before the termination of a history (either
by the weight cutoff, absorption, or exiting the domain) is therefore
bound to the number of stationary iterations required to converge to
the specified tolerance. In the case of the adjoint Neumann-Ulam
method, no such tolerance exists, however, we have specified a weight
cutoff $W_c$ under which low-weight histories will be prematurely
terminated as their contributions are deemed minute. After $k$
iterations, a stationary method is terminated as the error has reached
some fraction, $\epsilon$, of the initial error:
\begin{equation}
  ||\ve{e}^{k}||_2 = \epsilon ||\ve{e}^0||_2\:.
  \label{eq:linear_k_iter_norm4}
\end{equation}
Per Eq~(\ref{eq:linear_k_iter_norm3}), we see that this fraction is
equivalent to $\epsilon = \rho(\ve{H})^k$. For the adjoint Neumann-Ulam
method, if we take this fraction to be the weight cutoff, a measure of
how accurately the contributions of a particular history to the
solution are tallied, we then have the following empirical relation
for $k$:
\begin{equation}
  k = \frac{ \log(W_c) }{ \log( \rho(\ve{H}) ) }\:.
  \label{eq:analytic_k}
\end{equation}
This then gives us a means to estimate the length of the random walks
that will be generated from a particular linear operator based on the
Eigenvalues of its iteration matrix (independent of the linear
operator splitting chosen) and based on the weight cutoff parameter
used in the Neumann-Ulam method

\subsection{Wigner Rational Approximation}
In a domain decomposed situtation, not all histories will remain
within the domain they started in and must instead be
communicated. This communication, expected to be expensive, was
analyzed by Siegel and colleagues for idealized, load balanced
situations for full-core Monte Carlo\cite{siegel_analysis_2012}.  To
quantify the number of particles that leak out of the local domain
they define a leakage fraction, $\lambda$, as:
\begin{equation}
  \lambda = \frac{average\ \#\ of\ particles\ leaving\ local\ domain}
          {total\ of\ \#\ of\ particles\ starting\ in\ local\ domain}\:.
  \label{eq:leakage_fraction}
\end{equation}
For their studies, they assumed that the value of $\lambda$ was
dependent on the total cross section of the system via the Wigner
rational approximation. This approximation was developed by Wigner to
compute the neutron leakage probability from as a simple interpolation
between the small and large fuel lump cases, approximately
characterizing the leakage behavior of many common geometries.
\cite{duderstadt_nuclear_1976}. Outlined more thoroughly by Hwang's
chapter in \cite{azmy_nuclear_2010}, we begin the derivation of
Wigner's approximation by defining $l$ to be the average chord length
across the domain, and $\Sigma_t$ to be the total cross section in the
domain. We then bound the behavior for the leakage fraction in terms
of the optical thickness, $\Sigma_t l$, of the local domain. In the
case of a small domain or a small cross section (and therefore small
probabilty of absorption or scatter), $\Sigma_t l \rightarrow 0,
\lambda \rightarrow 1$. For a large domain or large cross section,
$\Sigma_t l \rightarrow \infty, \lambda \rightarrow (\Sigma_t
l)^{-1}$, where $(\Sigma_t l)^{-1}$ is effectively the number of
interactions that will occur over the length of the coord. Using
Wigner's simple interpolation, we then have a simple formula for the
domain leakage, $\lambda = 1/(1+\Sigma_t l)$. This provides a leakage
estimate, valid over the entire range of $\Sigma_t l$, in terms of the
optical thicknes of the system.

In the case of domain decomposed linear operator equations, the chord
length can be interpreted as the number of discrete states spanning a
domain, $n_i$, along the typical length of that domain (e.g. the
number of discrete states in the $i$ direction in a square domain
using our model diffusion problem). We then seek the equivalent of the
total cross section, $\Sigma_t$ in terms of the discrete random walk
behavior. The total cross section can also be interpreted as the
inverse of the mean free path of particles in the system. In the
Neumann-Ulam method, this is analagous to the length of the Markov
chains (and therefore the number of state-to-state transitions)
generated while sampling the Neumann series. In the previous section,
we determined this length with Eq~(\ref{eq:analytic_k}) to be
dependent on the Eigenvalues of the operator. In addition, we must
also consider the number of states that may be traversed in a given
transition, $n_s$, along the path of the typical length of the
system. For our diffusion model problem, $n_s$ would equate to the
number of states in the $i$ (or $j$ as the problem is symmetric)
direction that a history could potentially move and is dependent on
the stencil used for the discretization. We then have the following
relation for the \textit{optical thickness}, $\tau$, of a linear
operator equation domain:
\begin{equation}
  \tau = \frac{n_i n_s}{k}\:,
  \label{eq:matrix_optical_thickness}
\end{equation}
and inserting Eq~(\ref{eq:analytic_k}) gives:
\begin{equation}
  \tau = \frac{n_i n_s \log(\rho(\ve{H}))}{\log(W_c)}\:.
  \label{eq:matrix_optical_thickness_2}
\end{equation}
Using this optical thickness, we can then complete the Wigner rational
approximation by defining the bounds of $\tau \rightarrow 0, \lambda
\rightarrow 1$ and $\tau \rightarrow \infty, \lambda \rightarrow
\tau^{-1}$. We expect this behavior as optically thin domains will
communicate most of their histories while optically thick domains will
leak the fraction of histories that did not interact within. Using
the Wigner approximation, we can then define the leakage fraction out
of a domain for the adjoint Neumann-Ulam method:
\begin{equation}
  \lambda = \frac{1}{1+\tau}\:.
  \label{eq:analytic_domain_leakage}
\end{equation}
Here the fraction is explictly bound to the Eigenvalues of the
iteration matrix, the size of the domain, the content of the
discretization stencil, and the weight cutoff selected to terminate
low weight histories.  We also note here an interesting consequence of
this approximation: no matter how optically thick a domain may be, the
domain is still expected to leak some histories and thus communicate
them. Therefore, no matter how large the domains are made, if they do
not encapsulate the entire global domain they require some
communication in order to implement a domain decomposed adjoint
Neumann-Ulam method.

@@

%%---------------------------------------------------------------------------%%
\section{Conclusion}
These relationships will serve as guidelines for
selecting an appropriate parallel algorithm strategy and provide a
basis for performance models for future parallel implementations of
the adjoint Neumann-Ulam method.
